Type-Safe Modular Parsing

### 2017-04-24 SLE_17/sections/Discussion.tex
Copy content

\begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::=  ... \alt `\\' <ident> `:' <type> `.' <expr>
\end{grammar}
\end{tabular}

@@

\begin{figure}[t]
\begin{tabular}{m{0.42\linewidth}m{0.52\linewidth}}
\lstinputlisting[linerange=113-132]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
&
\lstinputlisting[linerange=136-156]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\end{tabular}
\caption{Language components \lstinline{VarExpr} (left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

>>>>

\huang{todo}
%% \begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
%% \setlength{\grammarindent}{5em}
%% \begin{grammar}
%% <type> ::= `int' \alt <type> `->' <type>
%% \end{grammar}
%% &
%% \setlength{\grammarindent}{5em}
%% \begin{grammar}
%% <expr> ::=  ... \alt `\\' <ident> `:' <type> `.' <expr>
%% \end{grammar}
%% \end{tabular}

@@

\huang{todo}
%% \begin{figure}[t]
%% \begin{tabular}{m{0.42\linewidth}m{0.52\linewidth}}
%% \lstinputlisting[linerange=113-132]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
%% &
%% \lstinputlisting[linerange=136-156]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
%% \end{tabular}
%% \caption{Language components \lstinline{VarExpr} (left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
%% \end{figure}

### 2017-04-24 SLE_17/sections/Introduction.tex
Copy content

\begin{comment}
A reason why a solution to the Expression Problem is important in
practice is that it is necessary for the development of
\emph{Software-Product Lines} (SPLs)~\cite{}. A software-product line
is a reusable set of components, which can be combined in multiple ways
to obtain different programs. Programming languages offer a concrete
example for SPLs. A SPL for programming languages would allow us to
model various typical operations of programming languages (such as
evaluation, compilation, or parsing) for various different language
constructs (such as binding, arithmetic, conditional or loops)
independently and separately. For example, evaluation components could be defined
independently for binding and arithmetic constructs. If the language
to be implemented is the pure lambda calculus, only evaluation of
binding constructs is necessary. However, more realistic programming
languages will include arithmetic constructs, and will require
evaluation for such constructs as well. In this case
both the component for evaluation of binders and arithmetic
expressions can be combined to implement the desired functionality.
\end{comment}

\begin{comment}
Most programming languages share alot of features in
common.

For example, most languages have language constructs for:
binding (such as variables, functions, and function applications);
basic arithmetic operations; basic logic and conditional operations;
loops; as well as various other features. For each language construct,
various operations (such as evaluation, compilation, or parsing) need
to be implemented. It is reasonable to wonder whether we can simply
implement those features independently of a particular implementation
of a programming language. Evaluation could be defined independently
for binding and arithmetic constructs. If the language to be
implemented is the pure lambda calculus, only evaluation of binding
constructs is necessary. Thus only the component that implements
evaluation for binding needs to be used in such an implementation.
However, more realistic programming languages
will include arithmetic constructs, and will require an evaluation
function for those.

Then it would be possible to \emph{reuse}
some of those features in \emph{multiple} different implementations of
programming languages. Essentially, this would enable a SPL for
programming languages, where all

A solution to the Expression Problem could ena

A concrete
example that illustrates this issue is
\end{comment}

@@

\begin{comment}
  By analising the \emph{full} grammar it is possible to remove
  backtracking, which would otherwise increase parsing times. Many
  parsing combinator libraries routinely use backtracting elimination
  to achieve performance. However, in a modular setting this technique
  cannot be used, because the full grammar is not known. Thus we have
  to be very conservative at eliminating backtracting. Unfortunatelly,
  this has a severe impact on performance.
\end{comment}

@@

\begin{itemize}[noitemsep,nolistsep]

\item {{\bf A Technique for Modular Parsing:}} We present a technique
that allows the development of semantically modular parsers.
The technique relies on the combination of Packrat parsing, multiple inheritance
  and Object Algebras.

\item {{\bf A Parsing Technique for OO ASTs:}} A simplified version of
  our technique also enables parsing OO-style ASTs, where new language
  constructs can be easily added.

\item {{\bf A Methodology for Writing Modular Parsing Code:}} We
  identify possible pitfalls using parser combinators. To avoid such
  pitfalls, we propose guidelines for writing parsing code using
  \emph{left-recursion} and \emph{longest-match composition}.

\item {{\bf TAPL case study:}} We conduct a case study with 18 interpreters
  from the TAPL book. The case study show the effectiveness of modular
  parsing in terms of reuse. The TAPL case study is available online
  at:
  \url{http://}\footnote{{\bf Note to Reviewers:} Please find the URL in the
  supplementary material for the submission.}

>>>>

@@

@@

\begin{itemize}

\item \textbf{A Technique for Modular Parsing:}
  We present a technique that allows the development of semantically modular parsers.
  The technique relies on the combination of Packrat parsing, multiple inheritance
  and Object Algebras.

\item \textbf{A Parsing Technique for OO ASTs:} A simplified version of
  our technique also enables parsing OO-style ASTs, where new language
  constructs can be easily added.

\item \textbf{A Methodology for Writing Modular Parsers:} We
  identify possible pitfalls using parser combinators. To avoid such
  pitfalls, we propose guidelines for writing parsing code using
  \emph{left-recursion} and \emph{longest-match composition}.

\item \textbf{TAPL case study:} We conduct a case study with 18 interpreters
  from the TAPL book. The case study show the effectiveness of modular
  parsing in terms of reuse. The TAPL case study is available online
  at:
  \url{http://}\footnote{\textbf{Note to Reviewers:} Please find the URL in the
    supplementary material for the submission.}

### 2017-04-24 SLE_17/sections/ObjectAlgebras.tex
Copy content

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser has type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

@@

\paragraph{A Solution}
A solution to this problem is instead have a field with the Object Algebra
in \lstinline{ExprOAParser}. This approach supports overriding as well.
The parsing code is as follows:

>>>>

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

@@

\paragraph{A Solution}
A solution to this problem is instead having a field with the Object Algebra
in \lstinline{ExprOAParser}. This approach supports overriding as well.
The parsing code is as follows:

### 2017-04-24 SLE_17/sections/OpenRecursion.tex
Copy content

\lstinputlisting[linerange=8-12]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PVAR

@@

\lstinputlisting[linerange=16-17]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PAPP

@@

\lstinputlisting[linerange=29-32]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PAPP2

@@

\lstinputlisting[linerange=44-49]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PLIT
\bruno{END: Text to be deleted/moved/rewritten in S2}

@@

\lstinputlisting[linerange=63-63]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PAPPFIX

@@

\lstinputlisting[linerange=67-75]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_COMBINATOR

@@

\lstinputlisting[linerange=79-79]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PVARAPP

@@

\lstinputlisting[linerange=83-102]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_TYPEDEXPR

>>>>

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PVAR

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PAPP

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PAPP2

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PLIT
\bruno{END: Text to be deleted/moved/rewritten in S2}

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PAPPFIX

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_COMBINATOR

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_PVARAPP

@@

\lstinputlisting[linerange=-]{}% APPLY:linerange=PARSING_TYPEDEXPR

### 2017-04-24 SLE_17/sections/PackratParsing.tex
Copy content

\item {\bf Modular parsers should be left-recursive;}

\item {\bf Modular parsers should use a longest match composition operator.}

@@

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseInt
\end{lstlisting}
\end{tabular}

@@

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseSub <|> parseInt
parseSub = do
  e <- parseExpr
  ...
\end{lstlisting}
\end{tabular}

@@

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `-' <int> <expr'>
\end{grammar}
&
\end{tabular}

@@

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
def ~[U](q: => Parser[U]): Parser[~[T, U]]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition. \\
\hline
\begin{lstlisting}
def ^^[U](f: (T) => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for function application. \\
\hline
\begin{lstlisting}
def ^^^[U](v: => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
\hline
\begin{lstlisting}
def <~[U](q: => Parser[U]): Parser[T]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
\hline
\begin{lstlisting}
def ~>[U](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
\hline
\begin{lstlisting}
def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
\end{lstlisting} \\
\hspace{.2in}- A parser generator for interleaved repetitions. \\
\hline
\begin{lstlisting}
def ident: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches an identifier. \\
\hline
\begin{lstlisting}
def numericLit: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches a numeric literal. \\
\hline
\begin{lstlisting}
def |[U >: T](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative composition. \\
\hline
\begin{lstlisting}
def |||[U >: T](q0: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}

>>>>

\item \textbf{Modular parsers should be left-recursive;}

\item \textbf{Modular parsers should use a longest match composition operator.}

@@

\huang{todo}

%% \begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
%% \setlength{\grammarindent}{5em}
%% \begin{grammar}
%% <expr> ::= <int>
%% \end{grammar}
%% &
%% \begin{lstlisting}[language=PlainCode]
%% parseExpr = parseInt
%% \end{lstlisting}
%% \end{tabular}

@@

\huang{todo}
%% \begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
%% \setlength{\grammarindent}{5em}
%% \begin{grammar}
%% <expr> ::= <int> \alt <expr> `-' <int>
%% \end{grammar}
%% &
%% \begin{lstlisting}[language=PlainCode]
%% parseExpr = parseSub <|> parseInt
%% parseSub = do
%%   e <- parseExpr
%%   ...
%% \end{lstlisting}
%% \end{tabular}

@@

\huang{todo}
%% \begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
%% \setlength{\grammarindent}{5em}
%% \begin{grammar}
%% <expr> ::= <int> <expr'>

%% <expr'> ::= <empty> \alt `-' <int> <expr'>
%% \end{grammar}
%% &
%% \end{tabular}

@@

\huang{todo}

%% \begin{table}[t]
%% \begin{tabular}{l}
%% \hline
%% \begin{lstlisting}
%% def ~[U](q: => Parser[U]): Parser[~[T, U]]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for sequential composition. \\
%% \hline
%% \begin{lstlisting}
%% def ^^[U](f: (T) => U): Parser[U]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for function application. \\
%% \hline
%% \begin{lstlisting}
%% def ^^^[U](v: => U): Parser[U]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
%% \hline
%% \begin{lstlisting}
%% def <~[U](q: => Parser[U]): Parser[T]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
%% \hline
%% \begin{lstlisting}
%% def ~>[U](q: => Parser[U]): Parser[U]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
%% \hline
%% \begin{lstlisting}
%% def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser generator for interleaved repetitions. \\
%% \hline
%% \begin{lstlisting}
%% def ident: Parser[String]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser which matches an identifier. \\
%% \hline
%% \begin{lstlisting}
%% def numericLit: Parser[String]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser which matches a numeric literal. \\
%% \hline
%% \begin{lstlisting}
%% def |[U >: T](q: => Parser[U]): Parser[U]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for alternative composition. \\
%% \hline
%% \begin{lstlisting}
%% def |||[U >: T](q0: => Parser[U]): Parser[U]
%% \end{lstlisting} \\
%% \hspace{.2in}- A parser combinator for alternative with longest match composition. \\
%% \hline \\
%% \end{tabular}
%% \caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
%% \end{table}
